\vspace{1.0cm} \LARGE \bf MERS-CoV spillover at the camel-human interface

### 2017-11-21 mers-structure.tex

MERS-CoV may join the list of pathogens able to jump species barriers but not spread efficiently in the new host, but every system is distinct.
Pathogens such as \textit{Bacillus anthracis}, Andes hantavirus \citep{martinez_person--person_2005}, monkeypox \citep{reed_detection_2004}, triple reassortant and H3N2v influenza A viruses \citep{shinde_triple-reassortant_2009,epperson_human_2013} belong to this list and yet only triple reassortant viruses eventually contributed to a pandemic, due to epidemiology and phylodynamics confined to influenza A virus.
When information about a system is lacking either because it is recent or entirely novel or if the system is inherently unpredictable, sequence data can and should be used jointly with whatever data sources are available to provide the unique pathogen perspective on an infectious disease outbreak.
Although MERS-CoV seems to cause self-limiting outbreaks in humans with no evidence of worsening outbreaks over time, sustained evolution of the virus in the reservoir and flow of viral lineages into humans provides the substrate for a more transmissible variant of MERS-CoV to possibly emerge.
Previous research \citep{antia_role_2003} suggests that zoonotic pathogens become disproportionately more likely to evolve towards $R_{0}>1.0$ if $R_{0}$ is close to $1.0$ in the new host to begin with, which leaves MERS-CoV in a precarious position: on one hand our estimate of $R_{0}$ at $\approx 0.7$ makes MERS-CoV a potential pandemic hazard, on the other nothing is known of fitness landscapes or adaptive potential of MERS-CoV.
We encourage continued genomic surveillance of MERS-CoV in the camel reservoir and from sporadic human cases.

>>>>

Pathogens \textit{Bacillus anthracis}, Andes hantavirus \citep{martinez_person--person_2005}, monkeypox \citep{reed_detection_2004} and influenza A are able to jump species barriers but only influenza A viruses have historically resulted in pandemics \citep{lipsitch_viral_2016}.
MERS-CoV may join the list of pathogens able to jump species barriers but not spread efficiently in the new host.
Since its emergence in 2012, MERS-CoV has caused self-limiting outbreaks in humans with no evidence of worsening outbreaks over time.
However, sustained evolution of the virus in the reservoir and continued flow of viral lineages into humans provides the substrate for a more transmissible variant of MERS-CoV to possibly emerge.
Previous modeling studies \citep{antia_role_2003, park_multiple_2013} suggest a positive relationship between initial $R_{0}$ in the human host and probability of evolutionary emergence of a novel strain which passes the supercritical threshold of $R_{0}>1.0$.
This leaves MERS-CoV in a precarious position; on one hand its current $R_{0}$ of $\sim$0.7 is certainly less than 1, but its proximity to the supercritical threshold raises alarm.
With very little known about the fitness landscape or adaptive potential of MERS-CoV, it is incredibly challenging to predict the likelihood of the emergence more transmissible variants.
In light of these difficulties, we encourage continued genomic surveillance of MERS-CoV in the camel reservoir and from sporadic human cases to rapidly identify a supercritical variant, if one does emerge.
